---
title: "CdGF_aucrf"
author: "Nicholas Lesniak"
date: "September 15, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr', 'cowplot','dplyr','AUCRF')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

## Can Day 0 Community members predict C. difficile CFU?
```{r fig.height= 10, fig.width=10}

meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# select only mice with 431 cdiff
meta_file <- meta_file[meta_file$cdiff_strain == '431', ]

# subset shared
#Create df with relative abundances
stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]), ]
# use %in% instead of == because to ignore NA
# rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
rownames(stool_shared) <- meta_file$mouse_id[rownames(meta_file) %in% 
                                               rownames(stool_shared) ]
rel_abund_stool <- 100 * stool_shared / unique(apply(stool_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]

#random forest regression model to determine Day 0 OTU correlated with increased CFU
seed <- 1
n_trees <- 2001
iters <- 100

```

##Can we predict moribundity with the community structure at Day 0?

```{r fig.width=6, fig.height=5, results= 'hide'}
#create dataframe for RF-classification
meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0)) 

Predict_df <- merge(meta_file[meta_file$day %in% 0, 
                              c('cage_id', 'mouse_id', 'Euth_Early')],
                    rel_abund_stool, by.x = 'mouse_id', by.y = 'row.names')

Predict_early_euth_df <- select(Predict_df, -cage_id, -mouse_id)

# create RF model
set.seed(seed)
rf_early_aucrf <- AUCRF(Euth_Early ~ ., data = Predict_early_euth_df,
                              ntree = n_trees, pdel = 0.05, ranking = 'MDA')
summary(rf_early_aucrf)
otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
set.seed(seed)
random_roc <- roc(sample(c(1,0), length(otu_euth_probs[ , 2]), replace = T) ~ otu_euth_probs[ , 2])

aucrf_data <- Predict_early_euth_df[,c('Euth_Early',rf_early_aucrf$Xopt)]

#### 10-fold Cross validated - 10 repetitions
cv10f_aucs <- c()
cv10f_all_resp <- c()
cv10f_all_pred <- c()
for(j in 1:iters){
  set.seed(j)
  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
  cv10f_probs <- rep(NA,nrow(aucrf_data))
  for(i in seq(1,nrow(aucrf_data),5)){
    train <- aucrf_data[sampling[-(i:(i+4))],]
    test <- aucrf_data[sampling[i:(i+4)],]
    set.seed(seed)
    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
  }
  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
  cv10f_aucs[j] <- cv10f_roc$auc
}
cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)

#### leave-one-out
LOO_probs <- NULL
LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
 train <- aucrf_data[-i,]
  test <- aucrf_data[i,]
  set.seed(seed)
  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
})
LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)

#### leave-one-cage-out
LOcO_df <- Predict_df %>% 
                select(Euth_Early, cage_id, one_of(rf_early_aucrf$Xopt))
cage_list <- as.character(unique(LOcO_df$cage_id))

LOcO_probs <- c()
for(i in 1:length(cage_list)){
  by_cage <- LOcO_df$cage_id == cage_list[i]
  LOcO_predict <- select(LOcO_df, -cage_id)
  train <- LOcO_predict[!by_cage,]
  test <- LOcO_predict[by_cage,]
  set.seed(seed)
  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
  LOcO_probs[LOcO_df$cage_id == cage_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
}
LOcO_roc <- roc(LOcO_df$Euth_Early~LOcO_probs)

LOcO_cv10f_aucs <- c()
LOcO_cv10f_all_resp <- c()
LOcO_cv10f_all_pred <- c()
for(j in 1:iters){
  set.seed(j)
  shuffle <- runif(nrow(LOcO_df))
  LOcO_df_r <- LOcO_df[order(shuffle), ]
#  sampling <- sample(1:nrow(LOcO_df),nrow(LOcO_df),replace=F)
  LOcO_cv10f_probs <- rep(NA,nrow(LOcO_df_r))
  for(i in 1:length(cage_list)){
    by_cage <- LOcO_df_r$cage_id == cage_list[i]
    LOcO_predict <- select(LOcO_df, -cage_id)
    train <- LOcO_predict[-by_cage,]
    test <- LOcO_predict[by_cage,]
    set.seed(seed)
    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
    LOcO_cv10f_probs[by_cage] <- predict(temp_model$RFopt, test, type='prob')[,2]
  }
  LOcO_cv10f_roc <- roc(LOcO_df_r$Euth_Early~LOcO_cv10f_probs)
  LOcO_cv10f_all_pred <- c(LOcO_cv10f_all_pred, LOcO_cv10f_probs)
  LOcO_cv10f_all_resp <- c(LOcO_cv10f_all_resp, LOcO_df_r$Euth_Early)
  LOcO_cv10f_aucs[j] <- LOcO_cv10f_roc$auc
}
LOcO_cv10f_roc <- roc(LOcO_cv10f_all_resp~LOcO_cv10f_all_pred)


#### Random - 10-fold Cross validated - 10 repetitions

set.seed(seed)
ran_outcome_data <- cbind(Euth_Early = factor(sample(Predict_early_euth_df$Euth_Early,  
                                                     nrow(Predict_early_euth_df))), 
                             Predict_early_euth_df[,rf_early_aucrf$Xopt]) 
predict_random <-  predict(rf_early_aucrf$RFopt, ran_outcome_data, type = 'prob')[,2]
predict_random_roc <- roc(ran_outcome_data$Euth_Early ~ predict_random)
 
#### plot rocs
par(mar=c(4,4,1,1))
plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
mtext(side=2, text="Sensitivity", line=2.5)
mtext(side=1, text="Specificity", line=2.5)
legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_euth_roc$auc),
                               sprintf('Leave-1 mouse-out (AUC: %.3g)',LOO_roc$auc),
                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
),lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','purple','gray','red'), bty='n')
```


```{r fig.width=10, fig.height=8}
layout(matrix(c(1,1,1,1,4,4,4,
                2,2,2,5,3,3,3,
                2,2,2,5,3,3,3), nrow=3, byrow = TRUE))


#AUCRF curve
par(mar=c(5,5,1.5,1))
plot(rf_early_aucrf, ylim = c(0.7, 1.1))
#mtext('A', at=-90, font=2, side=3)

#Importance Plot
euth_otus <- rf_early_aucrf$Xopt
euth_tax <- get_tax(1, euth_otus, tax_file)
euth_tax$tax_label <- paste(euth_tax$tax, ' (', 
                            gsub('tu0*', 'TU ', rownames(euth_tax)), 
                                   ')', sep = '')
dotchart(rev(rf_early_aucrf$ranking[euth_otus]), labels=rev(euth_tax[euth_otus, 'tax_label']), xlab='Mean Decrease Accuracy')
#mtext('B', at=-0.014, font=2, side=3)

#Abundance stripchart or most predictive otus
euth_otus <- rev(rf_early_aucrf$Xopt)
early_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==0, euth_otus]/10000 + 1e-4
late_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==1, euth_otus]/10000 + 1e-4

par(mar=c(4, 9, 1, 1))
plot(1, type="n", ylim=c(0,length(euth_otus)*2), xlim=c(1e-4,1e-2), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n")
index <- 1
for(i in euth_otus){
  stripchart(at=index-0.35, jitter(early_abunds[,i], amount=1e-5), pch=21, bg="royalblue1", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
  stripchart(at=index+0.35, jitter(late_abunds[,i], amount=1e-5), pch=21, bg="orange", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
  segments(mean(early_abunds[,i]),index-0.7,mean(early_abunds[,i]),index, lwd=3)
  segments(mean(late_abunds[,i]),index+0.7,mean(late_abunds[,i]),index, lwd=3)
  index <- index + 2
}
axis(2, at=seq(1,index-2,2), labels=euth_tax[euth_otus, 'tax_label'], las=1, line=-0.5, tick=F, cex.axis=0.8)
axis(1, at=c(1e-4, 1e-3, 1e-2), label=c("0", "0.1", "1"))
legend('topright', legend=c("Mild", "Severe"), pch=c(21, 21), pt.bg=c("orange","royalblue1"), cex=0.7)
#mtext('C', at=1e-7, font=2, side=3)
```


Would adding initial increase/slope of CFU or day 1 CFU help predict?  

* Adding CFU decreases AUC and improves Random model, likely due to samples lost due to lack of cfu data which are biased towards the severe samples
    - Day 1 CFU loses 2 samples  [AUCRF model with Day 1 CFU](~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/CdGF_AUCRF_withCFU_day1.html)
    - Day 2 cfu loses 14 samples, 10 are severe  [AUCRF model with Day 2 CFU](~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/CdGF_AUCRF_withCFU_day2.html)

Would it help to add community information such as diversity or enterotype?

* When adding enterotype, diversity, and CFU, invsimpson and cfu appear as features, but when cfu is removed, invsimpson is no longer a predictor

But how to show since model is already approaching 1?  

* Rerun AUCRF with added features and see if they replace any OTUs

***  
***  

### Validate AUCRF with Randomized Input


```{r Random_validation, fig.width=6, fig.height=5, results= 'hide'}

# boruta feature selection
#set.seed(seed)
#Boruta(Euth_Early ~ ., data = Validate_early_euth_df, maxRuns = 200)
```


### Add ranges to RF ROC ??

```{r error_plot, eval=F}

caretmodel <- function(x) {
  fitControl <- trainControl(method = "repeatedcv",
    number = 10,
    repeats = 10,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = TRUE)
  model <- train(Euth_Early ~ ., data=x, 
                 trControl=fitControl, method="rf", metric="ROC")
  return(model)
}
test <- aucrf_data
test$Euth_Early <- factor(ifelse(test[ , 'Euth_Early'] == 1, 'Late', 'Early'))
random <- test
set.seed(seed)
random$Euth_Early <- sample(random$Euth_Early, nrow(random))
caret_aucrf <- caretmodel(test)
caret_random <- caretmodel(random)
caret_prediction <- caretmodel(Predict_early_euth_df)

str(aucrf_data)
head(caret_aucrf$pred)

roclobster <- ggplot(caret_aucrf$pred, aes(d = obs, m = Late)) +
  geom_roc(n.cuts = 0) +
  theme_classic() +
  theme(
    axis.line.x = element_line(colour = "black"),
    axis.line.y = element_line(colour = "black")
  ) +
  geom_segment(aes(x = 0, y = 0, xend = 1, yend = 1), linetype=2) +
  ylab("Sensitivity") +
xlab(paste("Inverse Specificity"))

confusionMatrix(data = caret_aucrf$pred[,'pred'], reference = caret_aucrf$pred[,'obs'])
confusionMatrix(data = caret_random$pred[,'pred'], reference = caret_random$pred[,'obs'])

n <- nrow(Predict_early_euth_df)
indextrain <- sample(1:n,round(0.6*n),replace=FALSE)
Xtrain     <- Predict_early_euth_df[ indextrain,-1]
Xtest      <- Predict_early_euth_df[-indextrain,-1]
Ytrain     <- Predict_early_euth_df[ indextrain,1]
Ytest      <- Predict_early_euth_df[-indextrain,1]

qrf <- quantregForest(x=Xtrain, y=Ytrain, nodesize=10,sampsize=30)
conditionalQuantiles  <- predict(qrf, Xtest, what=sd)
print(conditionalQuantiles[1:4,])



```

### Other prediction models

##### C50

```{r c50, eval=F}
install.packages('C50')
library(C50)

set.seed(seed)
shuffle <- runif(nrow(Predict_early_euth_df))
Predict_early_euth_df_r <- Predict_early_euth_df[order(shuffle), ]

c50_euth_model <- C5.0(Predict_early_euth_df_r[1:50 , -1], Predict_early_euth_df_r[1:50 , 1])
summary(c50_euth_model)
plot(c50_euth_model)
c50_euth_predict <- predict(c50_euth_model, Predict_early_euth_df_r[51:55, ])
table(Predict_early_euth_df_r[51:55, 1],c50_euth_predict)

mylibrary<-c("SL.glm","SL.randomForest","SL.svm","SL.glmnet") 
SL_euth_model <- SuperLearner(as.numeric(Predict_early_euth_df_r[ , 1]), Predict_early_euth_df_r[ , -1], SL.library = mylibrary)
predict(SL_euth_model, newdata = as.numeric(Predict_early_euth_df_r[ , 1]))

SL_euth_model$SL.predict
sum=0
tt<-length(Predict_early_euth_df_r)
for(i in 1:tt) {
sum<-sum+(Predict_early_euth_df_r[i,1]-SL_euth_model$SL.predict[i])^2
  } 
sumg <-sum/tt
sumg

listWrappers()

Remove_otu_df <- Predict_early_euth_df_r %>% 
  select(-Otu00074, -Otu00114, -Otu00050, -Otu00124, -Otu00120, -Otu00100)

set.seed(seed)
rf_rm_otu_early_aucrf <- AUCRF(Euth_Early ~ ., data = Remove_otu_df,
                              ntree = n_trees, pdel = 0.05, ranking = 'MDA')
rm_otu_euth_probs <- predict(rf_rm_otu_early_aucrf$RFopt, type = 'prob')
rm_otu_euth_roc <- roc(Remove_otu_df$Euth_Early ~ rm_otu_euth_probs[ , 2])

plot(rm_otu_euth_roc, col='gray', lwd=2, add=T, lty=1)
```