---
title: "CdGF_aucrf"
author: "Nicholas Lesniak"
date: "September 15, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

## Can Day 0 Community members predict C. difficile CFU?
```{r fig.height= 10, fig.width=10}

meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# select only mice with 431 cdiff
meta_file <- meta_file[meta_file$cdiff_strain == '431', ]

# subset shared
#Create df with relative abundances
stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]), ]
# use %in% instead of == because to ignore NA
# rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
rownames(stool_shared) <- meta_file$mouse_id[rownames(meta_file) %in% 
                                               rownames(stool_shared) ]
rel_abund_stool <- 100 * stool_shared / unique(apply(stool_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]

#random forest regression model to determine Day 0 OTU correlated with increased CFU
seed <- 1
n_trees <- 2001
iters <- 100

```

##Can we predict moribundity with the community structure at Day 0?

```{r fig.width=8, fig.height=8, results= 'hide'}
#create dataframe for RF-classification
meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0)) 

Predict_df <- merge(meta_file[meta_file$day %in% 0, 
                              c('cage_id', 'mouse_id', 'Euth_Early')],
                    rel_abund_stool, by.x = 'mouse_id', by.y = 'row.names')

Predict_early_euth_df <- select(Predict_df, -cage_id, -mouse_id)

# create RF model
set.seed(seed)
rf_early_aucrf <- AUCRF(Euth_Early ~ ., data = Predict_early_euth_df,
                              ntree = n_trees, pdel = 0.05, ranking = 'MDA')

otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])

aucrf_data <- Predict_early_euth_df[,c('Euth_Early',rf_early_aucrf$Xopt)]

#### 10-fold Cross validated - 10 repetitions
cv10f_aucs <- c()
cv10f_all_resp <- c()
cv10f_all_pred <- c()
for(j in 1:iters){
  set.seed(j)
  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
  cv10f_probs <- rep(NA,nrow(aucrf_data))
  for(i in seq(1,nrow(aucrf_data),5)){
    train <- aucrf_data[sampling[-(i:(i+4))],]
    test <- aucrf_data[sampling[i:(i+4)],]
    set.seed(seed)
    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
  }
  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
  cv10f_aucs[j] <- cv10f_roc$auc
}
cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)

#### leave-one-out
LOO_probs <- NULL
LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
 train <- aucrf_data[-i,]
  test <- aucrf_data[i,]
  set.seed(seed)
  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
})
LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)

#### leave-one-cage-out
LOcO_df <- Predict_df %>% 
                select(Euth_Early, cage_id, one_of(rf_early_aucrf$Xopt))
cage_list <- as.character(unique(LOcO_df$cage_id))

LOcO_probs <- c()
for(i in 1:length(cage_list)){
  by_cage <- LOcO_df$cage_id == cage_list[i]
  LOcO_predict <- select(LOcO_df, -cage_id)
  train <- LOcO_predict[!by_cage,]
  test <- LOcO_predict[by_cage,]
  set.seed(j)
  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
  LOcO_probs[LOcO_df$cage_id == cage_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
}
LOcO_roc <- roc(LOcO_df$Euth_Early~LOcO_probs)


#### plot rocs
par(mar=c(4,4,1,1))
plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
mtext(side=2, text="Sensitivity", line=2.5)
mtext(side=1, text="Specificity", line=2.5)
legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_euth_roc$auc),
                               sprintf('Leave-1 mouse-out (AUC: %.3g)',LOO_roc$auc),
                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
),lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','purple','gray','red'), bty='n')
```


```{r fig.width=10, fig.height=10}


#Importance Plot
euth_otus <- rf_early_aucrf$Xopt
euth_tax <- get_tax(1, euth_otus, tax_file)
euth_tax$tax_label <- paste(euth_tax$tax, ' (', 
                            gsub('tu0*', 'TU ', rownames(euth_tax)), 
                                   ') : ', round(rf_early_aucrf$ranking[euth_otus], 3), sep = '')


RF_features_polt <- Predict_early_euth_df %>% 
  select(Euth_Early, one_of(euth_otus)) %>% 
  gather('OTU', 'Value', -Euth_Early) %>% 
  full_join(data.frame(OTU = names(rf_early_aucrf$ranking[euth_otus]), 
               MDA =  rf_early_aucrf$ranking[euth_otus])) %>% 
  ggplot( aes(x = reorder(factor(OTU), MDA), y = Value, fill = Euth_Early, dodge = Euth_Early)) +
  theme(legend.title=element_blank(), legend.position = c(0.9,0.5),
        panel.background = element_rect(fill = NA),
        panel.grid.major = element_line(colour = NA),
        panel.grid.minor = element_line(colour = NA)) +
  geom_point(pch = 21, position = position_jitterdodge(), alpha = 0.4, size = 2) + coord_flip() +
  scale_x_discrete(breaks = rownames(euth_tax), labels = euth_tax$tax_label) +
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                 geom = "crossbar", width = 0.6, position=position_dodge(width=0.8)) + 
  scale_fill_manual(values = c('red', 'black'), labels = c('Severe','Mild'),
                    guide = guide_legend(reverse=TRUE))

RF_features_polt + labs( y = 'Relative Abundance (%)\nClassification (OTU) : MDA', x = '', title = '')

RF_features_polt + labs( y = 'Relative Abundance (%) - log10\nClassification (OTU) : MDA', x = '', title = '') +
  scale_y_log10()
```


Would adding initial increase/slope of CFU or day 1 CFU help predict?  

* Adding CFU decreases AUC and improves Random model, likely due to samples lost due to lack of cfu data which are biased towards the severe samples
    - Day 1 CFU loses 2 samples  [AUCRF model with Day 1 CFU](~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/CdGF_AUCRF_withCFU_day1.html)
    - Day 2 cfu loses 14 samples, 10 are severe  [AUCRF model with Day 2 CFU](~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/CdGF_AUCRF_withCFU_day2.html)

Would it help to add community information such as diversity or enterotype?

* When adding enterotype, diversity, and CFU, invsimpson and cfu appear as features, but when cfu is removed, invsimpson is no longer a predictor

But how to show since model is already approaching 1?  

* Rerun AUCRF with added features and see if they replace any OTUs  

Leave 1 cage out RF with all otus gives and AUC of 0.63-0.68, but using the reduced features selected by AUCRF on the whole dataset gives an AUC of 0.88-0.90  

***  
***  
