---
title: "humanCdGF_predict_cfu"
author: "Nicholas Lesniak"
date: "October 7, 2016"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=10, fig.height=5)
```

```{r load packages}
pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```


```{r setup environment}
pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

```{r setup data}
# read in files
meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T)
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, day, log_cfu) %>% 
  filter(!is.na(log_cfu))

shared_list_days <- select(meta_file, -log_cfu)

# create shared with relative abundances over 1%
mean_otu_counts <- unique(apply(shared_file, 1, sum))
rel_abund_shared <- 100 * shared_file / mean_otu_counts
rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
rel_abund_shared_01$sample_id <- rownames(rel_abund_shared_01)

seed <- 1
n_trees <- 2001
rf_plot_list <- list()
n_features <- 10
cutoff <- 30

heat.plot <- function(input_maxtrix, high_color = '#660000', low_color = '#FFFFFF', legend_title = NA) {
  long.df <- data.frame(input_maxtrix)
  colnames(long.df) <- c(1:10)
  long.df$community_day <- as.character(c(0:10))
  long.df %>% gather(cfu_day, variable, -community_day) %>% 
    ggplot(aes(cfu_day, community_day)) + 
    geom_tile(aes(fill = variable), color = 'white') +
    scale_fill_gradient(low = low_color, high = high_color, name = legend_title) +
    labs(x = 'Day (CFU)', y = 'Day (Community)') + 
    theme(axis.line.x = element_blank(), axis.line.y = element_blank(), axis.ticks = element_blank()) + 
    scale_x_discrete(expand = c(0, 0), limits = c('1','2','3','4','5','6','7','8','9','10')) + 
    scale_y_discrete(expand = c(0, 0), limits = c('0','1','2','3','4','5','6','7','8','9','10')) + 
    theme(legend.position = 'top', legend.key.width = unit(2, 'cm'))
}
```

```{r functions}
rsq <- matrix(NA, nrow = 11, ncol = 10)
mse <- matrix(NA, nrow = 11, ncol = 10)
otu_votes <- rep(0, ncol(rel_abund_shared_01))
names(otu_votes) <- colnames(rel_abund_shared_01)
importance.df <- c()
for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
    for(n in 1:10){
      cfu_shared.df <- meta_file %>% 
        filter(day == n) %>% 
        select(mouse_id, log_cfu) %>% 
        right_join(community.df) %>% 
        select(-mouse_id) %>% 
        filter(!is.na(log_cfu))

      set.seed(seed)
      rf_all <- randomForest(log_cfu ~ ., data = cfu_shared.df,
                           importance = TRUE, ntree = n_trees)
      importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
      top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
      set.seed(seed)
      rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
                                                                 names(importance_sorted)[1:n_features])], 
                                  importance=TRUE, ntree=n_trees)
      rsq[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
      mse[m+1, n] <- round(tail(rf_n_features$mse, 1), 2)
      otu_votes <- otu_votes + 1*(colnames(rel_abund_shared_01) %in% top_important_OTU)
      importance.df <- rbind(importance.df, c(RSQ = round(tail(rf_n_features$rsq, 1), 2), 
                      MSE = round(tail(rf_n_features$mse, 1), 2),
                      importance(rf_all)[,1]/max(importance(rf_all)[,1])))
    }
  }

importance.df <- data.frame(importance.df)
import_features <- importance.df[importance.df$RSQ >= 0.6 & importance.df$MSE <= 0.8, ]
import_features <- import_features[ , !colnames(import_features) %in% c('RSQ','MSE')]
med_import_feat <- apply(import_features, 2, median)
otus_above_20 <- import_features[,names(med_import_feat)[med_import_feat >= 0.2]]
top_n_overall <- names(otu_votes)[otu_votes >= cutoff]

rsq_n <- matrix(NA, nrow = 11, ncol = 10)
mse_n <- matrix(NA, nrow = 11, ncol = 10)
otu_votes_n <- rep(0, ncol(rel_abund_shared_01))
names(otu_votes_n) <- colnames(rel_abund_shared_01)

for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
  for(n in 1:10){
    cfu_shared.df <- meta_file %>% 
      filter(day == n) %>% 
      select(mouse_id, log_cfu) %>% 
      right_join(community.df) %>% 
      select(log_cfu, one_of(top_n_overall)) %>% 
      filter(!is.na(log_cfu))
    
    set.seed(seed)
    rf_all <- randomForest(log_cfu ~ ., data = cfu_shared.df,
                           importance = TRUE, ntree = n_trees)
    importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
    set.seed(seed)
    rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
                                                                       names(importance_sorted)[1:n_features])], 
                                  importance=TRUE, ntree=n_trees)
    rsq_n[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
    mse_n[m+1, n] <- round(tail(rf_n_features$mse, 1), 2)
    otu_votes_n <- otu_votes_n + 1*(colnames(rel_abund_shared_01) %in% top_important_OTU)
  }
}
taxa <- get_tax(1, names(otu_votes_n[otu_votes_n >= 1]), tax_file)
taxa_names <- paste0(taxa$tax, ' (', gsub('tu0*', 'TU ', rownames(taxa)),')')
```

### All OTUs input to Random Forest
```{r all otus, fig.width=10, fig.height=6}
ggdraw()+
  draw_plot(heat.plot(rsq, '#003300', legend_title = expression(paste('R'^'2'))), 0, 0, 0.5, 1) + 
  draw_plot(heat.plot(mse, 'white', '#000066', 'MSE'), 0.5, 0, 0.5, 1)
#plot(otu_votes[otu_votes >= 10], xlab = 'OTU', ylab = 'Freq')
```

### OTUs selected by >`r cutoff`% of models generated by all OTU models 
###### Using the following bacteria: `r taxa_names`

```{r top otus, fig.width=10, fig.height=6}

ggdraw()+
  draw_plot(heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2'))), 0, 0, 0.5, 1) + 
  draw_plot(heat.plot(mse_n, 'white', '#000066', 'MSE'), 0.5, 0, 0.5, 1)
#plot(otu_votes_n[otu_votes_n >= 1], xlab = 'OTU', ylab = 'Freq')
```

###### Re-runnning the program and only focusing on the OTUs predicting day 9/10 to see if there is consistence in the predictive features gives the following selection:  
Otu00004 Otu00005 Otu00015 Otu00019 Otu00030 Otu00199 Otu00200 Otu00250   
      14       17       13       15       13       16       17       19
      
```{r boruta feature selection, eval = F}
bor_all <- c()

for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
    for(n in 9:10){
      cfu_shared.df <- meta_file %>% 
        filter(day == n) %>% 
        select(mouse_id, log_cfu) %>% 
        right_join(community.df) %>% 
        select(-mouse_id) %>% 
        filter(!is.na(log_cfu))

      set.seed(seed)
      bor_all[[(m*2)+(n-8)]] <- Boruta(log_cfu ~ ., data = cfu_shared.df, maxRuns = 2000)
    }
  }

boruta.votes <- rep(0, ncol(rel_abund_shared_01))
names(boruta.votes) <- colnames(rel_abund_shared_01)
for(i in 1:22){
  boruta.votes <- boruta.votes + 1*(bor_all[[i]]$finalDecision   == 'Confirmed')
}
plot(boruta.votes)
boruta.votes[boruta.votes >= 16]

colnames(otus_above_20)
```

Boruta Confirmed the following OTUs as important for predicting day 9/10 cfu:   
Otu00004 Otu00005 Otu00015 Otu00030 Otu00081 Otu00199 Otu00200 Otu00250 Otu00297  
      21       22       21       21       21       22       22       22       20  

Selecting OTUs through collecting the features from the most predictive community/cfu models (R^2 >= 0.6 and MSE <= 0.8), then converting all % Increase in MSE to relative values and taking the median value of of each OTU, then selecting OTUs that fall above the median value results in the following OTUs:  
"Otu00001" "Otu00002" "Otu00004" "Otu00005" "Otu00012" "Otu00014" "Otu00015" "Otu00016" "Otu00019" "Otu00028" "Otu00030" "Otu00048" "Otu00081" "Otu00101" "Otu00118" "Otu00199" "Otu00200" "Otu00250" "Otu00297"
```{r }