---
title: "HumanCdGF_microbiota_predict_by_day"
author: "Nick Lesniak"
date: "August 18, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=10, fig.height=5)
```

```{r load packages}
pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```


```{r setup environment}

# read in files
meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T)
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# identify c difficile otus
cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(cdiff_strain == 431) %>% 
  select(mouse_id, cage_id, sample_id, day, cdiff_cfu, Early_Euth, log_cfu) %>% 
  filter(!is.na(log_cfu))

# create shared with relative abundances over 1%
mean_otu_counts <- unique(apply(shared_file, 1, sum))
rel_abund_shared <- 100 * shared_file / mean_otu_counts
rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
rel_abund_shared_01$sample_id <- rownames(rel_abund_shared_01)

# create df with day 0 communities with column of mouse_id
day_0_otu_by_mouse <- 
  meta_file %>% 
  filter(day == 0 ) %>% 
  select(sample_id, mouse_id) %>% 
  left_join(rel_abund_shared_01) %>% 
  select(-sample_id)
# create df with cfu merge to shared
cfu_by_day_comm <-
  meta_file %>% 
  filter(day %in% 1:10) %>% 
  select(sample_id, day, log_cfu) %>% 
  left_join(rel_abund_shared_01) %>% 
  select(-sample_id, -one_of(cdiff_otu))


seed <- 18
n_trees <- 2001
set.seed(seed)

rf_plot_list <- list()
n_features <- 10

RandomForest_by_day <- function(i, community_day_0){
  if(!is.logical(community_day_0)){
    stop('Input for community  - Need community_day_0 = T/F')
  } else {
    rf_data <- if(community_day_0 == TRUE){
      meta_file %>%
        filter(day == 1) %>% 
        select(log_cfu, mouse_id) %>%
        inner_join(day_0_otu_by_mouse) %>% 
        select(-mouse_id)
    } else {
      cfu_by_day_comm %>% 
        filter(day == 1) %>% 
        select(-day)
    } 

    rf_all <- randomForest(log_cfu ~ ., data = rf_data,
                   importance = TRUE, ntree = n_trees)
    importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
    rf_n_features <- randomForest(log_cfu ~ ., data=rf_data[ ,c( 'log_cfu',
                              names(importance_sorted)[1:n_features])], 
                           importance=TRUE, ntree=n_trees)

    predict_plot_day <- paste('prediction_day_', i, sep = '')
    importance_plot_day <- paste('importance_day_', i, sep = '')

    rf_prediction <- cbind(rf_data[, 'log_cfu'], 
                       data.frame(rf_n_features$predicted))
    colnames(rf_prediction) <- c('observed', 'predicted')

    predict_plot_day <-  
      ggplot(data = rf_prediction, aes(x=observed, y=predicted)) + geom_point() + 
        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + 
        theme_bw() + ggtitle(paste("Random Forest Prediction of C Diff CFU on Day ",
                                   i, sep = '')) +
        annotate('text', x=0.15*(max(rf_prediction$observed)), 
                 y=0.95*(max(rf_prediction$predicted)), 
                 label=paste('R^2 = ', round(tail(rf_n_features$rsq, 1), 2),'\nMSE = ', 
                             round(tail(rf_n_features$mse, 1), 2), sep = ''))

    top_important_OTU <- data.frame(head(importance_sorted, n_features))
    colnames(top_important_OTU) <- 'Importance'
    top_important_OTU$OTU <- rownames(top_important_OTU)
    otu_taxa <- get_tax(1, top_important_OTU$OTU, tax_file)

    importance_plot_day <- 
      ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
        geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU),
                                        labels = rev(paste(otu_taxa[,1],' (',
                                                       rownames(otu_taxa),')',
                                                       sep=''))) +
        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
  
    rf_plot_list[[i]] <- list(predict_plot_day,importance_plot_day)
  }
}
rf_day_0_comm_plots <- lapply(1:10, RandomForest_by_day, community_day_0 = T)
rf_by_day_com_plots <- lapply(1:10, RandomForest_by_day, community_day_0 = F)

```

Removing mice which were euthanized early did not significantly change the R2 or MSE

### Predict Day 1 CFU from Day 0 Community

```{r Day_1_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[1]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[1]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 1 CFU from Day 1 Community

```{r Day_1_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[1]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[1]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 2 CFU from Day 0 Community

```{r Day_2_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[2]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[2]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 2 CFU from Day 2 Community

```{r Day_2_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[2]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[2]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 3 CFU from Day 0 Community

```{r Day_3_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[3]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[3]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 3 CFU from Day 3 Community

```{r Day_3_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[3]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[3]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 4 CFU from Day 0 Community

```{r Day_4_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[4]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[4]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 4 CFU from Day 4 Community

```{r Day_4_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[4]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[4]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 5 CFU from Day 0 Community

```{r Day_5_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[5]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[5]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 5 CFU from Day 5 Community

```{r Day_5_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[5]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[5]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 6 CFU from Day 0 Community

```{r Day_6_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[1]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[1]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 6 CFU from Day 6 Community

```{r Day_6_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[6]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[6]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 7 CFU from Day 0 Community

```{r Day_7_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[7]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[7]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 7 CFU from Day 7 Community

```{r Day_7_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[7]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[7]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 8 CFU from Day 0 Community

```{r Day_8_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[8]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[8]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 8 CFU from Day 8 Community

```{r Day_8_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[8]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[8]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 9 CFU from Day 0 Community

```{r Day_9_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[9]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[9]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 9 CFU from Day 9 Community

```{r Day_9_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[9]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[9]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 10 CFU from Day 0 Community

```{r Day_10_CFU_Day_0_Community}
ggdraw()+
  draw_plot(rf_day_0_comm_plots[[10]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_day_0_comm_plots[[10]][[2]], 0.5, 0, 0.5, 1)

```

### Predict Day 10 CFU from Day 10 Community

```{r Day_10_CFU__Community}
ggdraw()+
  draw_plot(rf_by_day_com_plots[[10]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_by_day_com_plots[[10]][[2]], 0.5, 0, 0.5, 1)

```

***

Modeled with OTUs with max rel abund > 1% (216 OTUs) 
When modeling based on mean rel abund > 1% (22 OTUs), R^2 decreases ~0.1 and MSE increases ~ 1 and lose all coriobacteriaceae/Actinobacteria


***

### Otus associated with CFU C difficile and same day OTU

```{r fig.width=10, fig.height=10}
predictive_otus <- paste('Otu00', c('012','171','063','033','038',
                                    '062','078','007','209'),    # lost with max rel abund > 10
                                    #'007', '014', '035', '005'), # gain with max rel abund > 10
                         sep = '')
predictive_taxa <- get_tax(1, predictive_otus, tax_file)
predictive_taxa$otu <- rownames(predictive_taxa)
predictive_taxa$tax_label <- paste(predictive_taxa$tax, ' (', 
                                   gsub('tu0*', 'TU ', predictive_taxa$otu), 
                                   ')', sep = '')

cfu_by_day <-
  meta_file %>% 
  filter(day %in% 1:10) %>% 
  select(sample_id, day, log_cfu, Early_Euth) %>% 
  left_join(rel_abund_shared_01) %>% 
  select(-sample_id, -one_of(cdiff_otu))

cfu_by_day %>% 
  select(log_cfu, day, one_of(predictive_otus), Early_Euth) %>% 
  filter(!is.na(log_cfu)) %>% 
  gather('otu','abundance',contains('Otu')) %>% 
  full_join(predictive_taxa) %>% 
  ggplot(aes(x = abundance, y = log_cfu, color = factor(Early_Euth))) + geom_point() + 
    facet_wrap(~ tax_label, scales = 'free_x') +
    labs( x = 'Relative Abundance', y = 'CFU (log10)')

```

```{r fig.width=10, fig.height=10}
cfu_by_day %>% 
  select(log_cfu, day, one_of(predictive_otus), Early_Euth) %>% 
  filter(!is.na(log_cfu)) %>% 
  gather('otu','abundance',contains('Otu')) %>% 
  full_join(predictive_taxa) %>% 
  ggplot(aes(x = log10(abundance), y = log_cfu, color = factor(Early_Euth))) + geom_point() + 
    facet_wrap(~ tax_label, scales = 'free_x') +
    labs( x = 'Relative Abundance (log10)', y = 'CFU (log10)')

```

### Day 0 OTUs associated with C difficile CFU

```{r fig.width=10, fig.height=10}
meta_file %>% 
  select(day, log_cfu, mouse_id, Early_Euth) %>% 
  full_join(day_0_otu_by_mouse) %>% 
  select(day, log_cfu, one_of(predictive_otus), Early_Euth) %>% 
  filter(!day == 0) %>% 
  gather('otu', 'abundance', contains('Otu')) %>% 
  filter(!is.na(abundance)) %>% 
  full_join(predictive_taxa) %>% 
  ggplot(aes(x = abundance, y = log_cfu, color = factor(Early_Euth))) + geom_point() + 
    facet_wrap(~ tax_label, scales = 'free_x') +
    labs( x = 'Relative Abundance', y = 'CFU (log10)')

```

```{r fig.width=10, fig.height=10}
meta_file %>% 
  select(day, log_cfu, mouse_id, Early_Euth) %>% 
  full_join(day_0_otu_by_mouse) %>% 
  select(day, log_cfu, one_of(predictive_otus), Early_Euth) %>% 
  filter(!day == 0) %>% 
  gather('otu', 'abundance', contains('Otu')) %>% 
  filter(!is.na(abundance)) %>% 
  full_join(predictive_taxa) %>% 
  ggplot(aes(x = log10(abundance), y = log_cfu, color = factor(Early_Euth))) + geom_point() + 
    facet_wrap(~ tax_label, scales = 'free_x') +
    labs( x = 'Relative Abundance (log10)', y = 'CFU (log10)')

```


```{r fig.width=10, fig.height=10}

corr_df_day_0_comm <- meta_file %>% 
  select(day, log_cfu, mouse_id) %>% 
  full_join(day_0_otu_by_mouse) %>% 
  filter(!day == 0) %>% 
  select(-day, -mouse_id)

get_CFU_cor <- function(dataframe, CFU, OTU){
    cor.test(dataframe[ , CFU], dataframe[ , OTU], method="spearman")[c("estimate","p.value")]
}

otu_cfu_corrs <- sapply(2:ncol(corr_df_day_0_comm), get_CFU_cor, CFU = 1, dataframe = corr_df_day_0_comm)
colnames(otu_cfu_corrs) <- colnames(corr_df_day_0_comm)[-1]

# identify the significant correlations after applying a correction for multiple
# comparisons
sig <- which(p.adjust(otu_cfu_corrs["p.value",], method="BH") < 0.05)
sigg_corr_otus <- data.frame(estimate = unlist(otu_cfu_corrs["estimate", sig]),
                   p_value = unlist(otu_cfu_corrs["p.value", sig]))
sigg_corr_otus$otus <- gsub('.rho', '', rownames(sigg_corr_otus))
sig_corr_phyla <- get_tax(1, sigg_corr_otus$otus, tax_file)
sigg_corr_otus$phylogeny <- paste(sig_corr_phyla[,'tax'], ' (', 
                                  gsub('tu0*', 'TU ', sigg_corr_otus$otu), 
                                   ')', sep = '')

sigg_corr_otus %>%
  filter(estimate >= 0.3 | estimate <= -0.3) %>%
  mutate(color = ifelse(estimate >= 0.32 | estimate <= -0.32, 'black', 'red')) %>% 
  ggplot( aes(x = reorder(factor(phylogeny), estimate), y = estimate)) + 
  geom_bar(stat = 'identity', aes(fill = color)) + coord_flip() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(values = c(black = 'black', red = 'red'), guide = FALSE) +
  labs( x = '', y = 'rho', title = 'Significant Spearman Correlation Coefficient - CFU vs OTU\n(only rho > 0.3 or < -0.3 shown)')

  
# '199','026','050','015','028','033','135','059','044'),    # lost with max rel abund > 10

```

```{r}
# function to compare rel abund of otu with cdiff cfu
plot_otu_w_cfu <- function(OTU){
  rel_abund_otu <- data.frame(rel_abund = rel_abund_shared_01[,OTU],
                              sample_id = rownames(rel_abund_shared_01))
  plot <- meta_file %>% 
  select(sample_id, day, cage_id, log_cfu) %>% 
  left_join(rel_abund_otu) %>% 
  select(cage_id, day, log_cfu, rel_abund) %>% 
  filter(day %in% 0:10) %>%
  group_by(cage_id, day) %>% 
  summarise(abund = mean(rel_abund), cfu = mean(log_cfu)) %>% 
  gather(data_type, values, abund:cfu) %>% 
  ggplot(aes(x = factor(day), y = values, group = data_type, color = data_type)) +
            facet_wrap(~cage_id) + 
            geom_line() + scale_y_log10()
}            

```