---
title: "Predictions From Day 0 Community Structures"
author: "Nicholas Lesniak"
date: "July 13, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message=FALSE}
if(!require(randomForest)) install.packages('randomForest', repos = 'http://cran.us.r-project.org')
library(randomForest)
if(!require(miscTools)) install.packages('miscTools', repos = 'http://cran.us.r-project.org')
library(miscTools)
if(!require(ggplot2)) install.packages('ggplot2', repos = 'http://cran.us.r-project.org')
library(ggplot2)
```

```{r}
setwd("~/Documents/Github/Schubert_humanCdGF_XXXX_2016")

# read in files
meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 'group')
shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header = T, row.names='Group')

# add log transformed CFU
meta_file$log_cfu <- log10(meta_file$cdiff_cfu + 1)

# remove cecum samples
meta_file <- meta_file[! meta_file$sample_type == 'cecum', ]

# subset shared
#Create df with relative abundances
stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
# use %in% instead of == because to ignore NA
# rename day 0 communities by mouse id
rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% rownames(stool_shared) ]
rel_abund_stool <- 100*stool_shared/unique(apply(stool_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_stool <- rel_abund_stool[,OTUs_stool_0_01]
OTU_list <- colnames(rel_abund_stool)

#Merge OTU data with mouse/exp data
#Predict_df <- merge(rf_meta_df, rel_abund_stool, by.x='RowNames', by.y='row.names')
#Predict_df <- Predict_df[,!names(Predict_df) %in% c('Mouse_ID','RowNames','cage_id')]

#predict day cfu by day 0 community
predict_df <- merge(meta_file[meta_file$day==2, c('Mouse_ID', 'log_cfu') ], stool_shared, by='Mouse_ID', by.y='row.names', all.y = T)
predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]

RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu),], importance = TRUE)

varImpPlot(RF)
ggplot(data = data.frame(actual=predict_df[,'log_cfu'],pred=predict(RF,predict_df)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
percent_var_expl <- RF$rsq[length(RF$rsq)]*100

#create function for random forest
run_rf_on <- function(DATAFRAME, RESPONSE, SAMPLE_SPLIT = 2/3,  OTUs = OTU_list){
set.seed(2)
     #
     Predictor_df <- as.data.frame(DATAFRAME[!DATAFRAME[,RESPONSE] %in% NA,c(RESPONSE,OTUs)])
     sample_set <- sample(seq_len(nrow(Predictor_df)), size = floor((SAMPLE_SPLIT)*nrow(Predictor_df)))
     test <- Predictor_df[-sample_set,]
     train <- Predictor_df[sample_set,] 
          RF <- randomForest(as.formula(paste(RESPONSE, '~ .')), data = train, importance = TRUE)
     R2 <- rSquared(test[,RESPONSE], test[,RESPONSE] - predict(RF,test))
     MSE <- mean((test[,RESPONSE] - predict(RF,test))^2)
     IMPORT <- importance(RF)
     FOREST <- RF
     PLOT<- ggplot(data = data.frame(actual=test[,RESPONSE],pred=predict(RF,test)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
     ##MEAN_DECREASE <- varImpPlot(RF)
     return(list(rf=FOREST,r2=R2,mse=MSE,import=IMPORT,plot=PLOT))
}
```

##Can we predict the endpoint C. difficile CFU with the community structure at Day 0?
But endpoint has data from Day 2,3,10

Might be misleading since contains CFU from different data points
     
```{r}
Day_Euth_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_Euth_cdiff_cfu')
Day_Euth_cdiff_cfu_rf$plot
```

% Variance explained `r Day_Euth_cdiff_cfu_rf$rf$rsq[length(Day_Euth_cdiff_cfu_rf$rf$rsq)]*100`  
r^2^ = `r Day_Euth_cdiff_cfu_rf$r2`  
MSE = `r Day_Euth_cdiff_cfu_rf$mse`  

***

##Can we predict moribundity with the community structure at Day 0?
```{r}
##Fix to be categorical, and not regression
Predict_early_euth_df <- Predict_df[!Predict_df$Euth_Early %in% NA,c('Euth_Early',OTU_list)]
rf_early_euth <- randomForest(as.factor(Euth_Early) ~ ., data = Predict_early_euth_df)
rf_early_euth
```

##Can we predict the Day 10 C. difficile CFU with the community structure at Day 0?

```{r}
Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
Day_10_cdiff_cfu_rf$plot
```

% Variance explained `r Day_10_cdiff_cfu_rf$rf$rsq[length(Day_10_cdiff_cfu_rf$rf$rsq)]*100`

Why day 10? What is significant about day 10?

***

##Does Day 0 community structure predict the each day C. difficile CFU at a similar?

##Can we predict the Day 1 C. difficile CFU with the community structure at Day 0?

```{r}
#rf to predict 
Day_1_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_1_cdiff_cfu')
Day_1_cdiff_cfu_rf$plot
```

% Variance explained `r Day_1_cdiff_cfu_rf$rf$rsq[length(Day_1_cdiff_cfu_rf$rf$rsq)]*100`

# Day 2 C. difficile CFU
```{r}
Day_2_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_2_cdiff_cfu')
Day_2_cdiff_cfu_rf$plot
```

% Variance explained `r Day_2_cdiff_cfu_rf$rf$rsq[length(Day_2_cdiff_cfu_rf$rf$rsq)]*100`

# Day 3 C. difficile CFU
```{r}
Day_3_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_3_cdiff_cfu')
Day_3_cdiff_cfu_rf$plot
```

% Variance explained `r Day_3_cdiff_cfu_rf$rf$rsq[length(Day_3_cdiff_cfu_rf$rf$rsq)]*100`

# Day 4 C. difficile CFU
```{r}
Day_4_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_4_cdiff_cfu')
Day_4_cdiff_cfu_rf$plot
```

% Variance explained `r Day_4_cdiff_cfu_rf$rf$rsq[length(Day_4_cdiff_cfu_rf$rf$rsq)]*100`

# Day 5 C. difficile CFU
```{r}
Day_5_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_5_cdiff_cfu')
Day_5_cdiff_cfu_rf$plot
```

% Variance explained `r Day_5_cdiff_cfu_rf$rf$rsq[length(Day_5_cdiff_cfu_rf$rf$rsq)]*100`

# Day 6 C. difficile CFU
```{r}
Day_6_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_6_cdiff_cfu')
Day_6_cdiff_cfu_rf$plot
```

% Variance explained `r Day_6_cdiff_cfu_rf$rf$rsq[length(Day_6_cdiff_cfu_rf$rf$rsq)]*100`

# Day 7 C. difficile CFU
```{r}
Day_7_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_7_cdiff_cfu')
Day_7_cdiff_cfu_rf$plot
```

% Variance explained `r Day_7_cdiff_cfu_rf$rf$rsq[length(Day_7_cdiff_cfu_rf$rf$rsq)]*100`

# Day 8 C. difficile CFU
```{r}
Day_8_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_8_cdiff_cfu')
Day_8_cdiff_cfu_rf$plot
```

% Variance explained `r Day_8_cdiff_cfu_rf$rf$rsq[length(Day_8_cdiff_cfu_rf$rf$rsq)]*100`

# Day 9 C. difficile CFU
```{r}
Day_9_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_9_cdiff_cfu')
Day_9_cdiff_cfu_rf$plot
```

% Variance explained `r Day_9_cdiff_cfu_rf$rf$rsq[length(Day_9_cdiff_cfu_rf$rf$rsq)]*100`

# Day 10 C. difficile CFU
```{r}
Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
Day_10_cdiff_cfu_rf$plot
```

% Variance explained `r Day_10_cdiff_cfu_rf$rf$rsq[length(Day_10_cdiff_cfu_rf$rf$rsq)]*100`


Predicting CFU from day 1 to 10 based on Day 0 community

There seems a flucuation in the % variance explained, which seems to coincide with the fluctuations in cdiff cfu  

## seed greatly affects variance, likely due to data split, sample size might be too small to make reliable predictions


```{r knitr::kable, warning=F, message=F}
var_by_day <- data.frame(Day=c(1:10),Percent_Variance_Explained=NA)
cdiff_cfu <- vector()
for (a in 1:10){
          var_by_day$Percent_Variance_Explained[a] <- eval(parse(text = paste('Day_', a, '_cdiff_cfu_rf$rf$rsq[length(Day_', a, '_cdiff_cfu_rf$rf$rsq)]*100', sep='')))
          cdiff_cfu <- c(cdiff_cfu,paste('Day_', a, '_cdiff_cfu', sep=''))
     }

knitr::kable(var_by_day)
plot(var_by_day, type='n', xlab='Day',ylab='Percent Var Explained')
lines(var_by_day$Day, var_by_day$Percent_Variance_Explained)

cfu_plot <- cbind(Mouse = paste(Predict_df$human_source,Predict_df$mouse_eartag, sep='_'),Predict_df[, cdiff_cfu])
names(cfu_plot) <- c('Mouse', c(1:10))
cfu_plot <- reshape2::melt(cfu_plot, by='Mouse')

ggplot(data=cfu_plot, aes(x=variable, y=value, group=Mouse)) + geom_line(aes(color=Mouse)) + geom_point() + scale_y_log10() +
     theme_classic() + theme(legend.position='none') + labs(title='C. Difficile CFU by mouse',x='Day',y='CFU')
```

***
***

##Can we predict day 1 weight change with the community structure at Day 0?
```{r}

```

##Can we predict day 10 weight change with the community structure at Day 0?
```{r}

```

##Can we predict ___________ with the community structure at Day 0?
```{r}

```
